Reducing Acceptance Marks in Emerson-Lei Automata by QBF Solving

### 2023-05-30 paper/main.tex
progress in final version

By choosing an appropriate acceptance condition, one can easily
represent many classical kinds of automata over infinite words. For
example, a transition-based B\"uchi automaton is a TELA with
$M = \{\tacc{1}{1}\}$, $\varphi = \Inf\tacc{1}{1}$, and the acceptance
mark \tacc{1}{1} being placed on all accepting transitions.
Similarly, a transition-based Rabin automaton with $k$ acceptance
pairs has
$M = \{\tacc{5}{1}, \tacc{3}{1'}, \ldots, \tacc{5}{k}, \tacc{3}{k'}\}$
and
$\varphi = (\Fin\tacc{5}{1} \land \Inf\tacc{3}{1'}) \lor \ldots \lor
(\Fin\tacc{5}{k} \land \Inf\tacc{3}{k'})$.  \fixme{A state-based acceptance
can be represented by TELA by placing the corresponding acceptance
marks on all transitions leaving the accepting states.}

@@

\end{itemize}
By $\vec{e},\vec{n},\vec{i},\vec{f}$ we denote the vectors of all
variables of the form $e_t$, $n_{t,k}$, $i_{c,k}$, and $f_{c,k}$,
respectively. The constructed QBF have the form
\[
  \Phi_{C,K}(\vec{n},\vec{i},\vec{f})~~=~~\forall \vec{e}~.~\mcycle(\vec{e})\implies\big(\msat_\varphi(\vec{e}) \iff \msat_{C,K}(\vec{e},\vec{n},\vec{i},\vec{f})\big),
\]
where $\forall\vec{e}$ denotes the sequence composed of $\forall e_t$
for all variables $e_t$ and the subformulas $\mcycle(\vec{e})$,
$\msat_\varphi(\vec{e})$, and
$\msat_{C,K}(\vec{e},\vec{n},\vec{i},\vec{f})$ are defined as
follows. \fixme{The subformula $\mcycle(\vec{e})$ will be defined as the last
one and we present three versions of it.}

@@

Let us now fix an SCC $S$. We define an equivalence
${\sim_S}\subseteq\delta_S\times\delta_S$ on transitions % in $\delta_S$
such that $t_1\sim_S t_2$ whenever $\mks(t_1)=\mks(t_2)$.
\begin{lemma} Assume that there is a function $\rem$ and a formula
  $\psi$ such that
  \[
    \textrm{for every set $\emptyset\neq T\subseteq\delta_S$ it holds }
      (T~\msat~\varphi~\iff~\rem(T)~\msat~\psi).\tag{$1$}\label{eq:lem1} 
  \]  
  Then there exists a function $\rem'$ that respects % the equivalence
  $\sim_S$ (i.e.,
  if $t_1\sim_S t_2$ then $\rem'(t_1)=\rem'(t_2)$) and
  \[
    \textrm{for every set $\emptyset\neq T\subseteq\delta_S$ it holds }
    (T~\msat~\varphi~\iff~\rem'(T)~\msat~\psi).\tag{$2$}\label{eq:lem2} 
  \]  
\end{lemma}
\begin{proof}
  Let $\rem$ be a function and $\psi$ a formula such that
  \eqref{eq:lem1} holds. To construct the function $\rem'$, we first
  select one transition from each equivalence class of~$\sim_S$. For
  every transition $t\in\delta_S$, by $\overline{t}$ we denote the
  selected transition equivalent to $t$ and we define the function
  $\rem'$ as $\rem'(t)=\rem(\overline{t})$. Note that we do not need
  to discuss the value of $\rem'$ on transitions outside $\delta_S$ as
  it is not relevant for the lemma. Clearly, $\rem'$ respects the
  equivalence $\sim_S$. It remains to show that \eqref{eq:lem2} holds
  for $\rem'$ and $\psi$.

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[smallautomaton]
    \begin{scope}[initial angle=90,xshift=-4cm]
      \node[state,initial] (p) at (0,0) {};
      \node[state] (q) at (1.8,0) {};
      \path[->,auto]
      (p) edge[out=-150,in=150,looseness=10] (p)
      (p) edge[out=30,in=150] (q)
      (q) edge[out=30,in=-30,looseness=10] (q)
      (q) edge[out=-150,in=-30] (p);
    \end{scope}

@@

    \begin{scope}[initial angle=90,densely dotted,black!35,xshift=4cm]
      \node[state,initial,solid,draw=black] (p) at (0,0) {};
      \node[state,solid,draw=black] (q) at (1.8,0) {};
      \path[->,auto]
      (p) edge[out=-150,in=150,looseness=10,black,solid] (p)
      (p) edge[out=30,in=150] (q)
      (q) edge[out=30,in=-30,looseness=10,black,solid] (q)
      (q) edge[out=-150,in=-30] (p);
    \end{scope}
  \end{tikzpicture}
  \caption{An automaton structure where the transition set forms a cycle
    (left) and two sets $\Te$ that
    are not cycles even if $\mcycle_2(\vec{e})$ holds (middle and
    right). The transition labels and acceptance marks are not depicted.}
  \label{fig:inter}
\end{figure}

\subsection{Strict version $\mcycle_3(\vec{e})$}
Before we give the definition of $\mcycle_3(\vec{e})$, we prove that
cycles can be characterised in the following way.
\begin{lemma}
  A nonempty set $T\subseteq\delta$ is a cycle if and only if for each
  set of states $S\subseteq Q$ one of the following conditions holds.
  \begin{enumerate}[A.]
  \item\label{en:1} All transitions in $T$ lead from a state in $S$ to a state in $S$ (i.e., $T\subseteq\delta_S$).
  \item\label{en:2} All transitions in $T$ lead from a state outside $S$ to a state outside $S$ (i.e., $T\subseteq\delta_{Q\smallsetminus S}$).
  \item\label{en:3} $T$ contains a transition leading from a state in $S$ to a state outside $S$ and a transition leading from a state outside
    $S$ to a state in $S$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  We first prove the direction ``$\Longrightarrow$''. Let $T$ be a
  cycle and $S\subseteq Q$ be an arbitrary set of states. We show that
  if (\ref{en:1}) and (\ref{en:2}) do not hold, then (\ref{en:3}) has
  to hold. Hence, assume that $T\not\subseteq\delta_S$ and
  $T\not\subseteq\delta_{Q\smallsetminus S}$. Then there are two
  cases.
  \begin{itemize}
  \item $T$ contains a transition $t\in\delta_S$ and a transition
    $t'\in\delta_{Q\smallsetminus S}$. The definition of a cycle
    implies that there exists a path $t_1t_2\ldots t_n\in T^+$ from a
    state $p$ back to $p$ containing both $t$ and $t'$. However, this
    implies that $T$ contains a transition leading from a state in $S$
    to a state outside $S$ and a transition leading from a state outside $S$
    to a state in $S$.
  \item $T$ contains a transition $t$ leading from a state in $S$ to a
    state outside $S$ (or vice versa). However, as $T$ is a cycle,
    there exists a path $t_1t_2\ldots t_n\in T^+$ that leads from a
    state $p$ to the same state and contains $t$. Hence, $T$ has to
    contain also a transition leading from a state outside $S$ to a
    state in $S$ (or vice versa).
  \end{itemize}
  In both cases, (\ref{en:3}) holds.

@@

\subsection{Complexity of formulas}
The constructed formulas $\Phi_{j,C,K}$ for $j\in\{1,2,3\}$ use
$|\delta|$ universally quantified variables $e_t$, $|\delta|\cdot K$
free variables $n_{t,k}$, and $C\cdot K$ free variables $i_{c,k}$ and
$f_{c,k}$. The formula $\Phi_{3,C,K}$ additionally uses $|Q|$
variables $s_q$ that are existentially quantified (when the formula is
transformed to prenex normal form).

@@

\subsection{Optimizations of formulas}

Finally, we mention three simple optimizations of the formula
construction, which are always applied in the rest of the paper.

@@

The presented reduction algorithms have been implemented in a tool
called \telatko. It is implemented in Python 3 and uses the Spot
library~\cite{duret.16.atva2} for automata parsing and manipulation,
and the theorem prover Z3~\cite{demoura.08.tacas} to solve the
satisfiability of QBF. Our tool is available at
\begin{center}
\url{https://gitlab.fi.muni.cz/xschwar3/telatko}
\end{center}
under the GNU GPLv3
license. The tool can be executed by the command 
\begin{center}
  \texttt{telatko -F <input.hoa> [-L }$j$\texttt{] [-C] [-I] [-T }$t$\texttt{] [-O <output.hoa>]}
\end{center}
where
\begin{description}
\item[\texttt{-F <input.hoa>}] specifies the file with the input automaton in HOA format~\cite{babiak.15.cav}, 
\item[\texttt{-L }$j$] specifies the reduction level; if omitted, the multi-level reduction is used,
\item[\texttt{-C}] switches on the reduction of the number of cubes
  after the number of marks is reduced (it corresponds to
  $\mathit{reduceC}=\true$ in Algorithms~\ref{alg:single}
  and~\ref{alg:multi}),
\item[\texttt{-I}] switches on the incremental version,
\item[\texttt{-T }$t$] sets the timeout for each QBF query to $t$ seconds (the default value is 50 seconds), 
\item[\texttt{-O <output.hoa>}] specifies the output file; if omitted,
  the produced automaton is sent to \textit{stdout} in the HOA format.
\end{description}

@@

%\todo{Upravit experimenty: pridat dalsi 2 sady automatu.}
\fixme{We have used the tool \texttt{genltl} of the Spot
library~\cite{duret.16.atva2} to generate LTL formulas from
literature, using all available formula sets implemented in
\texttt{genltl} 2.10.4.} The parametrized formulas were generated for all
combinations of parameter values from 1 to 4. We used the tool
\texttt{ltlfilt} of the Spot library to simplify the formulas and
remove duplicates. Altogether, we have generated 348 unique LTL
formulas. These have been translated to deterministic TELA by two
state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva}, namely \texttt{ltl2dela} (known as
Delag)~\cite{muller.17.gandalf} and \texttt{ltl2dgra} (known as
Rabinizer~4)~\cite{kretinsky.18.cav}, and to nondeterministic TELA by
another two state-of-the-art translators, namely \texttt{ltl2tgba}
from the Spot library~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}. Three of the considered
translators failed on some formulas: it usually reached a timeout of
60 seconds or it produced an automaton that cannot be parsed by the
Spot library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance marks as
there is a little point in reducing these. Table
\ref{tab:ltltranslators} shows the exact versions and homepages of
used translators. For each translator, it also provides the number of
fails, the number of produced automata with less than two marks, and
the number of automata with at least two marks. The last number is
typeset in bold as these automata are actually used for the
experimental evaluation of our reduction technique.

@@

Table~\ref{tab:cumm-data} shows the cumulative numbers of marks in
the input automata sets and after each reduction, together with the
reduction ratio and total time spent by the considered reduction. The
column \textit{solver timeout} shows the number of automata for which
the last query to QBF solver did not finish within the 30~seconds
limit. The timeout of the last QBF query implies that the automaton
may be potentially further reduced by the reduction with a longer time
limit. \fixme{The reduction ratio drop in higher levels, e.g.~between level~1
and level~2 for \texttt{ltl2dgra}, happens due to the QBF solver timeouts
occuring earlier as the formula being solved in higher level is more
complex. For \texttt{ltl2dela} and \texttt{ltl2dgra}, the automata sets
do not contain any automaton where level~2 or level~3 achieves a better
result than level~1. However, all levels contribute to the reductions in
the multilevel setting.}

@@

\begin{table}[b!]
\caption{The effect of reduction.
A cell on coordinates $(x,y)$ contains the number of automata that have been
reduced from $x$ to $y$ acceptance marks. If the cell contains a sum of two
numbers, the latter represents number of automata where the attempt to reduce
another mark has been unsuccessful due to a QBF solver timeout.}
\label{tab:heatmap}
\setlength{\tabcolsep}{2.5pt}
\centering
\input{table-heatmap.tex}
\end{table}

>>>>

By choosing an appropriate acceptance condition, one can easily
represent many classical kinds of automata over infinite words. For
example, a Büchi automaton can be represented as a TELA with the
acceptance condition $\varphi = \Inf\tacc{1}{1}$ and the single mark
$\tacc{1}{1}$ placed on all transitions leaving the accepting states
of the Büchi automaton. Further, a Rabin automaton with $k$ acceptance
pairs can be similarly represented as a TELA with acceptance condition
$\varphi = (\Fin\tacc{5}{1} \land \Inf\tacc{3}{1'}) \lor \ldots \lor
(\Fin\tacc{5}{k} \land \Inf\tacc{3}{k'})$ and marks
$M = \{\tacc{5}{1}, \tacc{3}{1'}, \ldots, \tacc{5}{k},
\tacc{3}{k'}\}$.

@@

\end{itemize}
By $\vec{e},\vec{n},\vec{i},\vec{f}$ we denote the vectors of all
variables of the form $e_t$, $n_{t,k}$, $i_{c,k}$, and $f_{c,k}$,
respectively. The constructed QBF have the form
\[
  \Phi_{C,K}(\vec{n},\vec{i},\vec{f})~~=~~\forall \vec{e}~.~\mcycle(\vec{e})\implies\big(\msat_\varphi(\vec{e}) \iff \msat_{C,K}(\vec{e},\vec{n},\vec{i},\vec{f})\big),
\]
where $\forall\vec{e}$ denotes the sequence composed of $\forall e_t$
for all variables $e_t$. Now we define the subformulas
$\msat_\varphi(\vec{e})$,
$\msat_{C,K}(\vec{e},\vec{n},\vec{i},\vec{f})$, and three versions of
$\mcycle(\vec{e})$.

@@

Let us now fix an SCC $S$. We define an equivalence
${\sim_S}\subseteq\delta_S\times\delta_S$ on transitions % in $\delta_S$
such that $t_1\sim_S t_2$ whenever $\mks(t_1)=\mks(t_2)$.
\begin{lemma} Assume that there is a function $\rem$ and a formula
  $\psi$ such that
  \[
    \textrm{for every set $\emptyset\neq T\subseteq\delta_S$ it holds }
      (T~\msat~\varphi~\iff~\rem(T)~\msat~\psi).\tag{$1$}\label{eq:lem1} 
  \]  
  Then there exists a function $\rem'$ that respects the equivalence
  $\sim_S$ (i.e., it assigns the same marks to equivalent transitions)
  % if $t_1\sim_S t_2$ then $\mks(\rem'(t_1))=\mks(\rem'(t_2))$)
  and
  \[
    \textrm{for every set $\emptyset\neq T\subseteq\delta_S$ it holds }
    (T~\msat~\varphi~\iff~\rem'(T)~\msat~\psi).\tag{$2$}\label{eq:lem2} 
  \]  
\end{lemma}
\begin{proof}
  Let $\rem$ be a function and $\psi$ a formula such that
  \eqref{eq:lem1} holds. To construct the function $\rem'$, we first
  select one transition from each equivalence class of~$\sim_S$. For
  every transition $t=(p,a,M',q)\in\delta_S$, by $\overline{t}$ we
  denote the selected transition equivalent to $t$ and we define the
  function $\rem'$ such that
  $\rem'(t)=(p,a,\mks(\rem(\overline{t})),q)$. Note that we do not need
  to discuss the value of $\rem'$ on transitions outside $\delta_S$ as
  it is not relevant for the lemma. Clearly, $\rem'$ respects the
  equivalence $\sim_S$. It remains to show that \eqref{eq:lem2} holds
  for $\rem'$ and $\psi$.

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[smallautomaton]
    \begin{scope}[initial angle=90,xshift=-4.5cm]
      \node[state,initial] (p) at (0,0) {};
      \node[state] (q) at (1.8,0) {};
      \path[->,auto]
      (p) edge[out=-150,in=150,looseness=10] (p)
      (p) edge[out=30,in=150] (q)
      (q) edge[out=30,in=-30,looseness=10] (q)
      (q) edge[out=-150,in=-30] (p);
    \end{scope}

@@

    \begin{scope}[initial angle=90,densely dotted,black!35,xshift=4.5cm]
      \node[state,initial,solid,draw=black] (p) at (0,0) {};
      \node[state,solid,draw=black] (q) at (1.8,0) {};
      \path[->,auto]
      (p) edge[out=-150,in=150,looseness=10,black,solid] (p)
      (p) edge[out=30,in=150] (q)
      (q) edge[out=30,in=-30,looseness=10,black,solid] (q)
      (q) edge[out=-150,in=-30] (p);
    \end{scope}
  \end{tikzpicture}
  \caption{An automaton structure % where the transition set forms a cycle
    (left) and two sets $\Te$ (middle and right) that are not
    cycles even if $\mcycle_2(\vec{e})$ holds. The transition labels
    and acceptance marks are not depicted.}
  \label{fig:inter}
\end{figure}

\subsection{Strict version $\mcycle_3(\vec{e})$}
Before we give the definition of $\mcycle_3(\vec{e})$, we prove that
cycles can be characterised in the following way.
\begin{lemma}
  A nonempty set $T\subseteq\delta$ is a cycle if and only if, for each
  set of states $S\subseteq Q$, one of the following conditions holds.
  \begin{enumerate}[A.]
  \item\label{en:1} All transitions in $T$ lead from a state in $S$ to a state in $S$ (i.e., $T\subseteq\delta_S$).
  \item\label{en:2} All transitions in $T$ lead from a state outside $S$ to a state outside $S$ (i.e., $T\subseteq\delta_{Q\smallsetminus S}$).
  \item\label{en:3} $T$ contains a transition leading from a state in $S$ to a state outside $S$ and a transition leading from a state outside
    $S$ to a state in $S$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  We first prove the direction ``$\Longrightarrow$''. Let $T$ be a
  cycle and $S\subseteq Q$ be an arbitrary set of states. We show that
  if (\ref{en:1}) and (\ref{en:2}) do not hold, then (\ref{en:3}) has
  to hold. Hence, assume that $T\not\subseteq\delta_S$ and
  $T\not\subseteq\delta_{Q\smallsetminus S}$. Then there are two
  cases.
  \begin{itemize}
  \item $T$ contains a transition $t\in\delta_S$ and a transition
    $t'\in\delta_{Q\smallsetminus S}$. The definition of a cycle
    implies that there exists a path $t_1t_2\ldots t_n\in T^+$ from a
    state $p$ back to $p$ containing both $t$ and $t'$. However, this
    implies that $T$ contains a transition leading from a state in $S$
    to a state outside $S$ and a transition leading from a state outside $S$
    to a state in $S$.
  \item $T$ contains a transition $t$ leading from a state in $S$ to a
    state outside $S$ (or vice versa). However, as $T$ is a cycle,
    there exists a path $t_1t_2\ldots t_n\in T^+$ that leads from a
    state $p$ to the same state and contains $t$. Hence, $T$ has to
    contain also a transition leading from a state outside $S$ to a
    state in $S$ (or vice versa).
  \end{itemize}
  In both cases, (\ref{en:3}) holds.

@@

\subsection{Complexity of formulas}
The constructed formulas $\Phi_{j,C,K}$ for $j\in\{1,2,3\}$ use
$|\delta|$ universally quantified variables $e_t$, $|\delta|\cdot K$
free variables $n_{t,k}$, and $C\cdot K$ free variables $i_{c,k}$ and
$f_{c,k}$. The formula $\Phi_{3,C,K}$ additionally uses $|Q|$
variables $s_q$ that are existentially quantified (when the formula is
transformed to prenex normal form) in the scope of universal
quantification of variables $e_t$.

@@

\subsection{Optimizations of formulas}
Finally, we mention three simple optimizations of the formula
construction, which are always applied in the rest of the paper.

@@

The presented reduction algorithms have been implemented in a tool
called \telatko. It is implemented in Python 3 and uses the Spot
library~\cite{duret.16.atva2} for automata parsing and manipulation,
and the theorem prover Z3~\cite{demoura.08.tacas} to solve the
satisfiability of QBF transformed to prenex normal form (and not using
CNF). Our tool is available at
\begin{center}
\url{https://gitlab.fi.muni.cz/xschwar3/telatko}
\end{center}
under the GNU GPLv3
license. The tool can be executed by the command 
\begin{center}
  \texttt{telatko -F <input.hoa> [-L }$j$\texttt{] [-C] [-I] [-T }$t$\texttt{] [-O <output.hoa>]}
\end{center}
where
\begin{description}
\item[\texttt{-F <input.hoa>}] specifies the file with the input automaton in HOA format~\cite{babiak.15.cav}, 
\item[\texttt{-L }$j$] specifies the reduction level; if omitted, the multi-level reduction is used,
\item[\texttt{-C}] switches on the reduction of the number of cubes
  after the number of marks is reduced (it corresponds to
  $\mathit{reduceC}=\true$ in Algorithms~\ref{alg:single}
  and~\ref{alg:multi}),
\item[\texttt{-I}] switches on the incremental version,
\item[\texttt{-T }$t$] sets the timeout for each QBF query to $t$ seconds (the default value is 50 seconds), 
\item[\texttt{-O <output.hoa>}] specifies the output file; if omitted,
  the produced automaton is sent to \textit{stdout} in the HOA format.
\end{description}

@@

To evaluate our reduction technique, we considered all sets of LTL
formulas from literature as they are provided by the tool
\texttt{genltl} of the Spot library~\cite{duret.16.atva2}~2.10.4, where
parametrized formulas were generated for all combinations of
parameter values from 1 to 4. We used the tool \texttt{ltlfilt} of the
Spot library to simplify the formulas and remove
duplicates. Altogether, we have generated 348 unique LTL
formulas. These have been translated to deterministic TELA by two
state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva}, namely \texttt{ltl2dela} (known as
Delag)~\cite{muller.17.gandalf} and \texttt{ltl2dgra} (known as
Rabinizer~4)~\cite{kretinsky.18.cav}, and to nondeterministic TELA by
another two state-of-the-art translators, namely \texttt{ltl2tgba}
from the Spot library~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}. Three of the considered
translators failed on some formulas: it usually reached a timeout of
60 seconds or it produced an automaton that cannot be parsed by the
Spot library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance marks as
there is a little point in reducing these. Table
\ref{tab:ltltranslators} shows the exact versions and homepages of
used translators. For each translator, it also provides the number of
fails, the number of produced automata with less than two marks, and
the number of automata with at least two marks. The last number is
typeset in bold as these automata are actually used for the
experimental evaluation of our reduction technique.

@@

Table~\ref{tab:cumm-data} shows the cumulative numbers of marks in the
input automata sets and after each reduction, together with the
reduction ratio and total time spent by the considered reduction. The
column \textit{solver timeout} shows the number of automata for which
the last query to QBF solver did not finish within the 30~seconds
limit. The timeout of the last QBF query means that the automaton may
be potentially further reduced if a longer time limit is used. One can
observe that a higher level sometimes achieves a smaller reduction
than a lower level (e.g., compare level~1 and level~2 for
\texttt{ltl2dgra}). This is caused by the QBF solver timeouts occuring
earlier as formulas constructed by the higher level are more
complex. For \texttt{ltl2dela} and \texttt{ltl2dgra}, the automata
sets do not contain any automaton where level~2 or level~3 achieves a
better result than level~1. However, all levels contribute to the
reductions in the multi-level setting.

@@

\begin{table}[b!]
\caption{The effect of multi-level reduction.
A cell on coordinates $(x,y)$ contains the number of automata that have been
reduced from $x$ to $y$ acceptance marks. If the cell contains a sum of two
numbers, the latter represents the number of automata where the attempt to reduce
another mark has been unsuccessful due to a QBF solver timeout.}
\label{tab:heatmap}
\setlength{\tabcolsep}{2.5pt}
\centering
\input{table-heatmap.tex}
\end{table}
